---
title: "transcriptomics_DgeUpsetPlot"
output: html_document
---
# transcriptomics_DgeUpsetPlot
This script compares significant genes from several differential gene expression (DGE) result tables and produces **UpSet plots plus export tables**.

**Expected input:** `.csv` DGE files in the current working directory, with the first column containing gene IDs and with `log2FoldChange` and `padj` columns present.

**Main outputs:** UpSet PDFs for upregulated, downregulated, separated up/down, and all significant genes, together with long/wide gene tables, presence matrices and exact-intersection tables.


```{r}
###############################################
# transcriptomics_DgeUpsetPlot -> simple DGE UpSet plots
#
# Outputs:
# - 1 PDF Upregulated
# - 1 PDF Downregulated
# - 1 Mixed PDF (up/down groups separated)
# - 1 All significant genes PDF (without separating up/down)
# - long tables
# - presence matrices
# - exact intersections
###############################################
```
The analysis logic below stays very close to the original script; the Markdown text simply makes each stage easier to follow.

## 0. Packages

Load the three packages used by the script. **Nothing to edit here** unless your package environment changes.

```{r}
# 0. Packages -------------------------------------------------------------

library(tidyverse)
library(ComplexUpset)
library(patchwork)
```

## 1. Parameters

These are the main values you may want to adjust. With the defaults, a gene is retained when **adjusted p-value < 0.05** and **|log2 fold change| ≥ 0.58**.

```{r}
# 1. Parameters -----------------------------------------------------------

padj_threshold <- 0.05
log2fc_threshold <- 0.58
max_intersections <- 50
pdf_width <- 13
pdf_height <- 8
```

## 2. Detect DGE files

The script searches the current directory for `.csv` files and uses the part of each filename before the first dot as the dataset name.

```{r}
# 2. List files --------------------------------------------------

dge_files <- list.files(pattern = "\\.csv$", full.names = TRUE)

if (length(dge_files) == 0) {
  stop("No .csv file found in the current directory.")
}

message("Files detected:")
message(paste(basename(dge_files), collapse = " | "))
```

## 3. Read and filter each DGE file

For each dataset, this function checks the expected columns, converts the relevant values to numeric format, applies the significance thresholds, and labels retained genes as **up** or **down**.

```{r}
# 3. Reading + filtering ---------------------------------------------------

read_and_filter_dge <- function(file,
                                padj_threshold = 0.05,
                                log2fc_threshold = 0.58) {

  sample_name <- sub("\\..*$", "", basename(file))

  df <- readr::read_csv(
    file,
    show_col_types = FALSE,
    col_types = cols(.default = col_character())
  )

  if (ncol(df) < 1) {
    stop(paste0("Empty or unreadable file: ", file))
  }

  names(df)[1] <- "gene"

  required_cols <- c("gene", "log2FoldChange", "padj")
  if (!all(required_cols %in% names(df))) {
    stop(paste0(
      "The file ", basename(file),
      " does not contain the required columns: ",
      paste(required_cols, collapse = ", "),
      "\nColumns present: ",
      paste(names(df), collapse = ", ")
    ))
  }

  df <- df %>%
    mutate(
      gene = as.character(gene),
      log2FoldChange = suppressWarnings(as.numeric(log2FoldChange)),
      padj = suppressWarnings(as.numeric(padj))
    )

  df_filt <- df %>%
    filter(
      !is.na(gene),
      gene != "",
      !is.na(padj),
      padj < padj_threshold,
      !is.na(log2FoldChange),
      abs(log2FoldChange) >= log2fc_threshold
    ) %>%
    mutate(
      sample = sample_name,
      direction = if_else(log2FoldChange > 0, "up", "down"),
      group = paste(
        sample,
        if_else(log2FoldChange > 0, "+", "-"),
        sep = "_"
      )
    ) %>%
    distinct(sample, direction, group, gene, .keep_all = TRUE)

  message("== ", basename(file), " ==")
  message("Significant genes retained: ", nrow(df_filt))
  message("Example gene IDs: ", paste(head(df_filt$gene, 10), collapse = " | "))

  df_filt
}
```

## 4. Combine all filtered datasets

Apply the filtering function to every detected CSV file, combine the results, and print a simple global check.

```{r}
# 4. Read all files -----------------------------------------------

dge_list <- lapply(
  dge_files,
  read_and_filter_dge,
  padj_threshold = padj_threshold,
  log2fc_threshold = log2fc_threshold
)

dge_all <- bind_rows(dge_list)

if (nrow(dge_all) == 0) {
  stop("No significant gene after filtering across all files.")
}

message("=== CHECK GLOBAL ===")
message("Total retained rows: ", nrow(dge_all))
message("Total unique genes: ", dplyr::n_distinct(dge_all$gene))
```

## 5. Utility functions

These functions prepare gene sets, presence matrices, export tables, exact intersections and the final UpSet plots. **For routine use, this section normally does not need editing.**

```{r}
# 5. Utilities ----------------------------------------------------------

make_sets <- function(df, grouping_col) {
  grouping_col <- rlang::ensym(grouping_col)
  group_name <- rlang::as_string(grouping_col)

  tmp <- df %>%
    group_by(!!grouping_col) %>%
    summarise(genes = list(sort(unique(gene))), .groups = "drop")

  out <- tmp$genes
  names(out) <- tmp[[group_name]]
  out
}

make_presence_df <- function(sets_list) {
  if (length(sets_list) == 0) {
    return(data.frame())
  }

  all_genes <- sort(unique(unlist(sets_list)))

  if (length(all_genes) == 0) {
    return(data.frame())
  }

  mat <- sapply(sets_list, function(gset) as.integer(all_genes %in% gset))

  if (is.null(dim(mat))) {
    mat <- matrix(mat, ncol = 1)
    colnames(mat) <- names(sets_list)
  }

  mat_df <- data.frame(
    gene = all_genes,
    mat,
    check.names = FALSE,
    stringsAsFactors = FALSE
  )

  mat_df
}

write_long_gene_table <- function(df, file) {
  out <- df %>%
    arrange(sample, direction, desc(log2FoldChange), gene) %>%
    select(sample, direction, group, gene, log2FoldChange, padj)

  write.csv(out, file = file, row.names = FALSE)
}

write_wide_padded_table <- function(df, grouping_col, file) {
  grouping_col <- rlang::ensym(grouping_col)

  group_values <- df %>%
    distinct(!!grouping_col) %>%
    pull(!!grouping_col)

  tables_list <- vector("list", length(group_values))
  names(tables_list) <- group_values

  for (g in group_values) {
    df_g <- df %>%
      filter((!!grouping_col) == g) %>%
      arrange(desc(log2FoldChange), gene) %>%
      select(gene, log2FoldChange, padj) %>%
      distinct(gene, .keep_all = TRUE)

    tables_list[[g]] <- df_g
  }

  if (length(tables_list) == 0) {
    write.csv(data.frame(), file = file, row.names = FALSE)
    return(invisible(NULL))
  }

  max_len <- max(vapply(tables_list, nrow, integer(1)))

  padded_tables <- lapply(names(tables_list), function(g) {
    df_g <- tables_list[[g]]
    n <- nrow(df_g)

    if (n < max_len) {
      df_g <- rbind(
        df_g,
        data.frame(
          gene = rep(NA_character_, max_len - n),
          log2FoldChange = rep(NA_real_, max_len - n),
          padj = rep(NA_real_, max_len - n)
        )
      )
    }

    colnames(df_g) <- c(
      paste0(g, "_gene"),
      paste0(g, "_log2FoldChange"),
      paste0(g, "_padj")
    )

    df_g
  })

  wide_table <- do.call(cbind, padded_tables)
  write.csv(wide_table, file = file, row.names = FALSE)
}

write_exact_intersections <- function(presence_df, file_prefix) {
  if (nrow(presence_df) == 0 || ncol(presence_df) <= 1) {
    write.csv(
      data.frame(),
      paste0(file_prefix, "_exact_intersections_summary.csv"),
      row.names = FALSE
    )
    write.csv(
      data.frame(),
      paste0(file_prefix, "_exact_intersections_genes.csv"),
      row.names = FALSE
    )
    return(invisible(NULL))
  }

  membership_df <- presence_df
  rownames(membership_df) <- membership_df$gene
  membership_df$gene <- NULL

  exact_labels <- apply(membership_df, 1, function(x) {
    present <- names(x)[x == 1]
    if (length(present) == 0) return(NA_character_)
    paste(sort(present), collapse = " & ")
  })

  genes_df <- data.frame(
    gene = rownames(membership_df),
    exact_intersection = exact_labels,
    stringsAsFactors = FALSE
  ) %>%
    filter(!is.na(exact_intersection)) %>%
    arrange(exact_intersection, gene)

  summary_df <- genes_df %>%
    count(exact_intersection, name = "n_genes") %>%
    arrange(desc(n_genes), exact_intersection)

  write.csv(
    summary_df,
    paste0(file_prefix, "_exact_intersections_summary.csv"),
    row.names = FALSE
  )

  write.csv(
    genes_df,
    paste0(file_prefix, "_exact_intersections_genes.csv"),
    row.names = FALSE
  )
}

make_upset_plot <- function(presence_df,
                            sets_vector,
                            title_text,
                            ylab_text,
                            max_intersections = 50,
                            base_size = 11) {

  if (nrow(presence_df) == 0) {
    stop("presence_df is empty.")
  }

  transparent_stripe <- grDevices::adjustcolor("white", alpha.f = 0)

  ComplexUpset::upset(
    presence_df,
    intersect = sets_vector,
    min_size = 1,
    n_intersections = max_intersections,
    width_ratio = 0.10,
    height_ratio = 0.70,
    stripes = ComplexUpset::upset_stripes(
      colors = c(transparent_stripe, transparent_stripe)
    ),
    name = "Intersection size",

    matrix = ComplexUpset::intersection_matrix() +
      ggplot2::theme(
        panel.border = ggplot2::element_rect(
          fill = NA,
          colour = "black",
          linewidth = 0.6
        ),
        panel.background = ggplot2::element_rect(
          fill = "white",
          colour = NA
        ),
        plot.background = ggplot2::element_rect(
          fill = "white",
          colour = NA
        )
      ),

    base_annotations = list(
      "Intersection size" = intersection_size(
        text = list(
          angle = 45,
          hjust = 0,
          vjust = -0.2,
          colour = "black",
          fontface = "bold",
          size = 4.5
        ),
        bar_number_threshold = 0,
        fill = "grey80",
        colour = NA
      )
    ),

    set_sizes = upset_set_size(
      geom = ggplot2::geom_bar(
        width = 0.35,
        fill = "grey80",
        colour = NA
      )
    ),

    themes = ComplexUpset::upset_modify_themes(list(
      "default" = ggplot2::theme(
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        strip.background = ggplot2::element_rect(fill = "white", colour = "black", linewidth = 0.6)
      ),
      "intersections_matrix" = ggplot2::theme(
        axis.text.x = ggplot2::element_blank(),
        axis.ticks.x = ggplot2::element_blank(),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.border = ggplot2::element_blank(),
        panel.grid.major = ggplot2::element_blank(),
        panel.grid.minor = ggplot2::element_blank(),
        strip.background = ggplot2::element_rect(fill = "white", colour = "black", linewidth = 0.6)
      ),
      "Intersection size" = ggplot2::theme(
        axis.text.x = ggplot2::element_blank(),
        axis.ticks.x = ggplot2::element_blank(),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.border = ggplot2::element_blank(),
        panel.grid.major = ggplot2::element_blank(),
        panel.grid.minor = ggplot2::element_blank(),
        strip.background = ggplot2::element_rect(fill = "white", colour = "black", linewidth = 0.6),
        strip.text = ggplot2::element_text(
          face = "bold",
          colour = "black",
          size = ggplot2::rel(1.75)
        )
      ),
      "overall_sizes" = ggplot2::theme(
        axis.text.x = ggplot2::element_text(
          angle = 45,
          hjust = 1,
          vjust = 1,
          face = "bold",
          colour = "black",
          size = ggplot2::rel(1.5)
        ),
        panel.background = ggplot2::element_rect(fill = "white", colour = NA),
        plot.background = ggplot2::element_rect(fill = "white", colour = NA),
        panel.border = ggplot2::element_blank(),
        panel.grid.major.x = ggplot2::element_line(colour = "grey80", linewidth = 0.3),
        panel.grid.major.y = ggplot2::element_blank(),
        panel.grid.minor.x = ggplot2::element_blank(),
        panel.grid.minor.y = ggplot2::element_blank(),
        strip.background = ggplot2::element_rect(fill = "white", colour = "black", linewidth = 0.6)
      )
    ))
  ) +
    ggplot2::labs(
      title = title_text,
      x = NULL,
      y = ylab_text
    ) +
    ggplot2::coord_cartesian(clip = "off") +
    ggplot2::theme_bw(base_size = base_size) +
    ggplot2::theme(
      text = ggplot2::element_text(colour = "black", face = "bold"),
      plot.title = ggplot2::element_text(
        face = "bold",
        colour = "black",
        size = ggplot2::rel(1.75)
      ),
      axis.title = ggplot2::element_text(face = "bold", colour = "black"),
      axis.text = ggplot2::element_text(face = "bold", colour = "black"),
      axis.text.x = ggplot2::element_blank(),
      axis.ticks.x = ggplot2::element_blank(),
      axis.text.y = ggplot2::element_text(face = "bold", colour = "black"),
      strip.text = ggplot2::element_text(face = "bold", colour = "black"),
      strip.background = ggplot2::element_rect(fill = "white", colour = "black", linewidth = 0.6),
      legend.title = ggplot2::element_text(face = "bold", colour = "black"),
      legend.text = ggplot2::element_text(face = "bold", colour = "black"),
      panel.background = ggplot2::element_rect(fill = "white", colour = NA),
      plot.background = ggplot2::element_rect(fill = "white", colour = NA),
      panel.border = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      panel.grid.major = ggplot2::element_blank(),
      plot.margin = ggplot2::margin(t = 10, r = 40, b = 10, l = 10)
    )
}
```

## 6. Build the comparison groups

Create four views of the same filtered DGE results: **upregulated**, **downregulated**, **up/down kept separate**, and **all significant genes merged by dataset**.

```{r}
# 6. Build groups ------------------------------------------------

up_sets <- dge_all %>%
  filter(direction == "up") %>%
  make_sets(sample)

down_sets <- dge_all %>%
  filter(direction == "down") %>%
  make_sets(sample)

mixed_sets <- dge_all %>%
  make_sets(group)

all_sets <- dge_all %>%
  make_sets(sample)

up_sets <- up_sets[lengths(up_sets) > 0]
down_sets <- down_sets[lengths(down_sets) > 0]
mixed_sets <- mixed_sets[lengths(mixed_sets) > 0]
all_sets <- all_sets[lengths(all_sets) > 0]

message("=== CHECK SETS ===")
message("Number of UP groups: ", length(up_sets))
message("Number of DOWN groups: ", length(down_sets))
message("Number of MIXED groups: ", length(mixed_sets))
message("Number of ALL groups: ", length(all_sets))

if (length(up_sets) > 0) {
  message("UP sizes: ", paste(names(up_sets), lengths(up_sets), sep = "=", collapse = " | "))
}
if (length(down_sets) > 0) {
  message("DOWN sizes: ", paste(names(down_sets), lengths(down_sets), sep = "=", collapse = " | "))
}
if (length(mixed_sets) > 0) {
  message("MIXED sizes: ", paste(names(mixed_sets), lengths(mixed_sets), sep = "=", collapse = " | "))
}
if (length(all_sets) > 0) {
  message("ALL sizes: ", paste(names(all_sets), lengths(all_sets), sep = "=", collapse = " | "))
}
```

## 7. Export tables

Write the filtered genes, presence matrices, exact intersections and group summaries to CSV files for inspection or downstream analysis.

```{r}
# 7. Tables ----------------------------------------------------------------

dge_up <- dge_all %>% filter(direction == "up")
dge_down <- dge_all %>% filter(direction == "down")

write.csv(
  dge_all %>%
    arrange(sample, direction, desc(log2FoldChange), gene) %>%
    select(sample, direction, group, gene, log2FoldChange, padj),
  file = "DGE_all_significant_genes_long.csv",
  row.names = FALSE
)

write_long_gene_table(dge_up, "Upregulated_genes_long.csv")
write_long_gene_table(dge_down, "Downregulated_genes_long.csv")
write_long_gene_table(dge_all, "Mixed_groups_genes_long.csv")
write_long_gene_table(dge_all, "All_significant_genes_long.csv")

write_wide_padded_table(dge_up, sample, "Upregulated_genes_by_dataset_wide.csv")
write_wide_padded_table(dge_down, sample, "Downregulated_genes_by_dataset_wide.csv")
write_wide_padded_table(dge_all, group, "Mixed_groups_genes_by_group_wide.csv")
write_wide_padded_table(dge_all, sample, "All_significant_genes_by_dataset_wide.csv")

up_presence <- make_presence_df(up_sets)
down_presence <- make_presence_df(down_sets)
mixed_presence <- make_presence_df(mixed_sets)
all_presence <- make_presence_df(all_sets)

write.csv(up_presence, "Upregulated_presence_matrix.csv", row.names = FALSE)
write.csv(down_presence, "Downregulated_presence_matrix.csv", row.names = FALSE)
write.csv(mixed_presence, "Mixed_groups_presence_matrix.csv", row.names = FALSE)
write.csv(all_presence, "All_significant_presence_matrix.csv", row.names = FALSE)

write_exact_intersections(up_presence, "Upregulated")
write_exact_intersections(down_presence, "Downregulated")
write_exact_intersections(mixed_presence, "Mixed_groups")
write_exact_intersections(all_presence, "All_significant")

summary_table <- bind_rows(
  tibble(group_type = "up_only", group = names(up_sets), n_genes = lengths(up_sets)),
  tibble(group_type = "down_only", group = names(down_sets), n_genes = lengths(down_sets)),
  tibble(group_type = "mixed", group = names(mixed_sets), n_genes = lengths(mixed_sets)),
  tibble(group_type = "all_significant", group = names(all_sets), n_genes = lengths(all_sets))
)

write.csv(summary_table, "UpSet_groups_summary.csv", row.names = FALSE)
```

## 8. Generate UpSet PDF figures

Generate the four UpSet plot families. Plot width is adjusted automatically to the number of sets, while `pdf_height` controls the height.

```{r}
# 8. PDF figures -----------------------------------------------------------

if (nrow(up_presence) > 0 && length(up_sets) > 0) {
  p_up <- make_upset_plot(
    presence_df = up_presence,
    sets_vector = names(up_sets),
    title_text = "Upregulated genes",
    ylab_text = "Genes per dataset",
    max_intersections = max_intersections
  )

  ggplot2::ggsave(
    filename = "UpSet_upregulated.pdf",
    plot = p_up,
    width = max(14, 1.5 * length(up_sets)),
    height = pdf_height,
    device = grDevices::cairo_pdf,
    limitsize = FALSE
  )
  message("PDF generated: UpSet_upregulated.pdf")
}

if (nrow(down_presence) > 0 && length(down_sets) > 0) {
  p_down <- make_upset_plot(
    presence_df = down_presence,
    sets_vector = names(down_sets),
    title_text = "Downregulated genes",
    ylab_text = "Genes per dataset",
    max_intersections = max_intersections
  )

  ggplot2::ggsave(
    filename = "UpSet_downregulated.pdf",
    plot = p_down,
    width = max(14, 1.5 * length(down_sets)),
    height = pdf_height,
    device = grDevices::cairo_pdf,
    limitsize = FALSE
  )
  message("PDF generated: UpSet_downregulated.pdf")
}

if (nrow(mixed_presence) > 0 && length(mixed_sets) > 0) {
  p_mixed <- make_upset_plot(
    presence_df = mixed_presence,
    sets_vector = names(mixed_sets),
    title_text = "All groups (Up/Down separated)",
    ylab_text = "Genes per group",
    max_intersections = max_intersections,
    base_size = 10
  )

  ggplot2::ggsave(
    filename = "UpSet_all_groups_mixed.pdf",
    plot = p_mixed,
    width = max(18, 1.35 * length(mixed_sets)),
    height = pdf_height,
    device = grDevices::cairo_pdf,
    limitsize = FALSE
  )
  message("PDF generated: UpSet_all_groups_mixed.pdf")
}

if (nrow(all_presence) > 0 && length(all_sets) > 0) {
  p_all <- make_upset_plot(
    presence_df = all_presence,
    sets_vector = names(all_sets),
    title_text = "All significant genes (Up/Down merged)",
    ylab_text = "Genes per dataset",
    max_intersections = max_intersections
  )

  ggplot2::ggsave(
    filename = "UpSet_all_significant_merged.pdf",
    plot = p_all,
    width = max(14, 1.5 * length(all_sets)),
    height = pdf_height,
    device = grDevices::cairo_pdf,
    limitsize = FALSE
  )
  message("PDF generated: UpSet_all_significant_merged.pdf")
}

message("Completed.")
```
